Bayesian Detection of Bias in Peremptory Challenges Using Historical Strike Data

### 2023-03-11 analysis/memo_TAS_reviewers.Rmd
minor edits; added comment on simulation results for alpha = 0

This memorandum replies to comments provided by the TAS associate editor and two reviewers. We thank the editor and reviewers for their comments. We believe the revised paper is much better as a result.

@@

1. As our revised Introduction now makes clear (p. 4-5 and Fig. 1), neither our statistical approach (nor any other) can alone detect _Batson_ violations, because _Batson_ procedure permits any type of admissible evidence of illegal strike bias -- not just the attorney's past strike data -- and requires the trial judge to weigh all such evidence to decide whether illegal strike bias exists.

@@

4. Absent some very strong assumptions, there is no generally justifiable hyper prior for $\alpha$ in this context of strike bias and _Batson_ challenges. This is because the proper weight to place on historical data depends on what causes incompatibility between the historical data and the current data. The plausible reasons for non-random missingness and such confounding, however, likely vary considerably depending on strike procedure, courtroom practices, prosecutor charging practices, among other things. As we now note in the Discussion section of the paper, the literature on the power prior discusses information criteria for guide values of $\alpha$ as well as treating $\alpha$ as random and assigning a prior to it. However, that literature is unclear as to what implies for _Batson_ and peremptory strikes, for at least three reasons.  First, unlike models of strike procedure discussed here, that literature has mostly focused on clinical trials and models whose properties are well known (e.g., the linear model). In the clinical trial setting, it is easier to measure trial-level (and patient-level) characteristics that may be responsible for incompatibility. Even in these settings, selecting a prior for $\alpha$ is not straightforward. @Neelon2010 warn that using random power parameters (in their case, with a beta prior distribution) tends to attenuate (more down-weighting) the influence of the historical data, and so one might have to use highly informative priors on the power parameter. These authors suggest that practitioners consider whether "it is more appropriate to assign $\alpha$ a fixed value based on expert opinion about the relevance of the historical data to the current analysis". More recent versions of the power prior are beginning to address this issue (e.g., using "commensurate power priors" [@Hobbs2011]). Second, our approach does not attempt to maximize coverage in the traditional sense, but aims only to "detect" bias by determining whether zero falls outside the chosen credible interval for the posterior distribution of the bias parameter. Third, whereas some papers treat incompatibility as itself a parameter [e.g., @Ollier2020], the precise approaches rely on strong assumptions that do not obviously make sense in our context.  Accordingly, we believe the most conservative choice (for now) is to keep $\alpha$ a fixed value and check for sensitivity of results to the choice of $\alpha$ value. Proceeding otherwise would require the work of a completely new paper.

@@

Fourth, the comment is overinclusive. Most work on the use of historical strike data (a) presupposes that _Batson_ law permits prior strikes as relevant evidence of bias in the current trial; (b) presupposes that any statistical approach therein cannot automatically determine _Batson_ violations alone; and (c) adjusts in some way for incompatibility between historical strike data and strikes from the current trial. As an example, we now point in the paper's Introduction to @Gastwirth2014. Our approach has a distinctive benefit of making incompatibility assumptions transparent via choosing a fixed value for the power prior's $\alpha$.

@@

Thank you for spotting this.  The paper has been revised accordingly.

@@

Thank you for spotting this.  The equation has been revised accordingly.

@@

We revised to remove such confusion. This includes changing the software app to use 95% credible intervals. On the choice of credible intervals, see Item 1(5) above. The revision does not necessarily fix the type I error, because this coverage probability refers to the true bias parameter (that is fixed in our simulations), whereas we are interested in "bias detection", i.e. whether the credible interval includes 0 or not. The paper also has a sensitivity analysis for the selection of the nominal credible interval level.

@@

This comment is not clear to us. The real data analyses are based on real historical strike data from real criminal cases in the federal district court in Connecticut. On how we collected this data, see the paper's Appendix. If they wish, users of the app can run the model excluding any historical data (as this reviewer suggested) by not selecting the name of a prosecutor or defense attorney. Like our model, the app uses the $N(0,2)$ prior for the bias parameter. Similarly, the app user can alter the weight of the historical strike data (if selected) by adjusting $\alpha$ (the power prior parameter). Thus, the app user can: (1) choose not to use historical data, (2) use historical data with equal weight as the current trial (alpha = 1), or (3) down-weight the historical trial data (alpha = 0.5 or alpha = 0.2). See also Item 1(6).

@@

This information has been added to the text (sec. 2.5, p. 18).

<!-- A normal distribution was assumed for the distribution of b. The length of the Markov chain was 110000 and the first 10000 iterations were dropped as burn-in. No thinning was performed, as the correlation was weak and the convergence occurred rapidly. Convergence was checked from both traceplots and the Gelman-Rubin statistic. Computation times varies from 0.76 to 1.2 seconds, depending in the amount of data used (current data only _versus_ current and historical data).-->

@@

Thank you for pointing out these issues. We have revised Equations (3) and (6) accordingly.

@@

Our statistical approach's validity turns neither on how many courts make available the kind of data it requires nor on proving how likely courts will start collecting such data. Indeed, as section 3.2 (paragraphs 1 and 2) of the paper explains, we hope our statistical approach can inform data collection efforts going forward both by courts and courtroom observers exercising their right at attend jury selection in criminal court proceedings.

@@

No, this is not correct. We have revised the paper and the software application to avoid such confusion. See Item 1(5) above. We agree that the choice of credible interval should be consistent within different stages of the _Batson_ challenge and across judges deciding _Batson_ challenges within the same jurisdiction.

@@

Thank you for this reference. The paper now cites to the model in Gastwirth (2005). On "data from an actual case," section 2.5 of the paper (our discussion about the prototype software application) expressly relies on actual strike data from a sample of criminal cases.

>>>>

This memorandum replies to comments from a TAS associate editor and two reviewers. We believe the revised paper is much better because of their comments.

@@

1. As our revised Introduction now makes clear (p. 4-5 and Figure 1), neither our statistical approach (nor any other) can alone detect _Batson_ violations, because _Batson_ procedure permits any type of admissible evidence of illegal strike bias -- not just the attorney's past strike data -- and requires the trial judge to weigh all such evidence to decide whether illegal strike bias exists.

@@

4. Absent some very strong assumptions, there is no generally justifiable hyper prior for $\alpha$ in this context of strike bias and _Batson_ challenges. This is because the proper weight to place on historical data depends on what causes incompatibility between the historical data and the current data, which in turn likely varies by. among other things, strike procedure and courtroom practices. As we now note in the Discussion section of the paper, the literature on the power prior discusses information criteria for guide values of $\alpha$ as well as treating $\alpha$ as random and assigning a prior to it. However, that literature is unclear as to what implies for _Batson_ and peremptory strikes, for at least three reasons. First, unlike models of strike procedure discussed here, that literature has mostly focused on clinical trials and models whose properties are well known (e.g., generalized linear models). In the clinical trial setting, it is easier to measure trial-level (and patient-level) characteristics that may be responsible for incompatibility. Even in these settings, selecting a prior for $\alpha$ is not straightforward. @Neelon2010 warn that using random power parameters (in their case, with a beta prior distribution) tends to attenuate (more down-weighting) the influence of the historical data, and so one might have to use highly informative priors on the power parameter. These authors suggest that practitioners consider whether "it is more appropriate to assign $\alpha$ a fixed value based on expert opinion about the relevance of the historical data to the current analysis." More recent versions of the power prior are beginning to address this issue (e.g., using "commensurate power priors" [@Hobbs2011]). Second, our approach does not attempt to maximize coverage in the traditional sense, but aims only to "detect" bias by determining whether zero falls outside the chosen credible interval for the posterior distribution of the bias parameter. Third, whereas some papers treat incompatibility as itself a parameter [e.g., @Ollier2020], the precise approaches rely on strong assumptions that do not obviously make sense in our context.  Accordingly, we believe the most conservative choice (for now) is to keep $\alpha$ a fixed value and check for sensitivity of results to the choice of $\alpha$ value. Proceeding otherwise would require the work of a completely new paper.

@@

Fourth, the comment is overinclusive. Most work on the use of historical strike data (a) presupposes that _Batson_ law permits prior strikes as relevant evidence of bias in the current trial; (b) presupposes that any statistical approach therein cannot automatically determine _Batson_ violations alone; and (c) adjusts in some way for incompatibility between historical strike data and strikes from the current trial. As an example, we now point in the paper's Introduction to @Gastwirth2014. Our approach has a distinctive benefit of making incompatibility assumptions transparent via choosing and justifying a fixed value for the power prior's $\alpha$.

@@

Thank you for spotting this.  We revised the paper accordingly.

@@

Thank you for spotting this.  We revised the equation accordingly.

@@

We revised to remove such confusion. This includes changing the prototype software app to use 95% credible intervals. On the choice of credible intervals, see Item 1(5) above. The revision does not necessarily fix the type I error, because this coverage probability refers to the true bias parameter (that is fixed in our simulations), whereas we are interested in "bias detection", i.e. whether the credible interval includes 0 or not. The paper also has a sensitivity analysis for the selection of the nominal credible interval level.

@@

This comment is not clear to us. We base the real data analyses on real historical strike data from real criminal cases in the federal district court in Connecticut. On how we collected this data, see the paper's Appendix. If they wish, users of the app can run the model excluding any historical data (as this reviewer suggested) by not selecting the name of a prosecutor or defense attorney. Like our model, the app uses the $N(0,2)$ prior for the bias parameter. Similarly, the app user can alter the weight of the historical strike data (if selected) by adjusting $\alpha$ (the power prior parameter). Thus, the app user can: (1) choose not to use historical data, (2) use historical data with equal weight as the current trial (alpha = 1), or (3) down-weight the historical trial data (alpha = 0.5 or alpha = 0.2). See also Item 1(6).

Similarly, although we did run simulations for when when $\alpha = 0$, Figures 2-4 do not depict the bias detection rate for that scenario, simply because that scenario is equivalent to evaluating bias detection with only strike data from the current trial (no historical strike data) -- a topic of prior papers [e.g. @Kadane2018a].

@@

We added this information to the paper (sec. 2.5, p. 18).

@@

Thank you for pointing out these issues. We revised Equations (3) and (6) accordingly.

@@

Our statistical approach's validity turns neither on how many courts make available the kind of data it requires nor on proving how likely courts will start collecting such data. Indeed, as section 3.2 (paragraphs 1 and 2) of the paper explains, we hope our statistical approach can inform data collection efforts going forward both by courts and courtroom observers exercising their right to attend jury selection in criminal court proceedings.

@@

No, this is not correct. We have revised the paper and the software application to avoid such confusion. See Item 1(5) above. We agree that the choice of credible interval should be consistent across different stages of a _Batson_ challenge and across judges deciding _Batson_ challenges within the same jurisdiction.

@@

Thank you for this reference. The paper now cites to the model in Gastwirth (2005). On "data from an actual case," section 2.5 of the paper (our discussion about the prototype software application) expressly relies on actual strike data from a sample of criminal cases. As mentioned above, we intend the reader to gain understanding by following along using our prototype software application.
